\ch{\textsc{RELIKE}}: Reionization Effective Likelihood from Planck 2018 Data (?)

### 2021-03-30 paper/rei.tex
Updates from Overleaf

\title{Effective Reionization Likelihood from Planck 2018 Data (?)}

@@

We publish Relike, an fast and accurate effective likelihood code based on constraints from the latest Planck 2018 data on the CMB reionization principal components (PC). This code allows one to constrain arbitrary global ionization history $x_e(z)$ between $6 < z < 30$ with Planck without having to modify CAMB each time a different model is tested. We tested the code on two example models
%: 1) the tanh model with a single transition redshift, and 2) a two-parameter model with two transitions allowing for a high-redshift ionization plateau. 
which showed good agreement with
%The Relike chains converged 300$\times$ and 100$\times$ faster for model 1 and 2 respectively, and showed good agreement compared to directly
sampling the exact Planck likelihoods.
We also find that contrary to the Planck 2015, the Planck 2018 PC posteriors are well approximated by a Gaussian in PC amplitude space, this allows us to replace the kernel density estimate (KDE) evaluation of the PC chains in the code with an even faster Gaussian evaluation. We supply both the KDE and Gausisan modes of the likelihood in a CosmoMC implementation, and the Gaussian mode alone in an easy-to-use standalone python likelihood package that can be connected to any other sampler. We expect the code release to dramatically reduce the time for testing individual ionization models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as 21cm, luminosity function, star formation history, etc.
%We expect the code release to be useful for individual model testing as the user would not need to modify CAMB every time a different ionization history model is needed; the effective likelihood code also makes it easier to perform joint constraints of the Planck data with other reionization data such as 21cm, luminosity functions, star formation rate measurements, etc.
Finally, we derive model-independent constraints: $\tau_{\rm PC} = ...$ and $\tau_{\rm PC}(15, 30) < XX $ (95\% C.L.). The upper limits on high-redshift optical depth is at least a factor of 2 looser those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method; we demonstrate robustness of our results by explicitly recovering  $\tau_{\rm PC}(15, 30) < XX $ (95\% C.L.) using a  two-parameter toy model without using PCs.

@@

The cosmic microwave background (CMB) has entered an era of precision cosmology. The measurements from the \textit{Planck} satellite have shown agreement with the standard $\Lambda$CDM model which describes the initial perturbations in the Universe and their evolution. While many components of the standard cosmology model is well-understood, the details of the process of reionization however, remains one of the most uncertain pieces. Its uncertainty propagates into the inferences of other important parameters such as the primordial power spectrum amplitude. Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB; and will also have implications for inferring cosmic acceleration through the growth of structure. \ch{[add more refs here]}

Typically, the impact of reionization on the primary CMB fluctuations has been modeled as a steplike transition in the global ionization history, with the step location parameterized by the total Thomson optical depth induced. This steplike model describes a Universe in which all hydrogen becomes fully ionized almost instantaneously at one particular redshift, and assumes, by construction, that there is negligible ionization before the transition. However, through the shape of the reionization bump induced in the CMB E-mode polarization at large angles, more information on the coarse-grained evolution of the ionization history can be obtained.

In fact, to extract the most information possible from this E-mode bump, Ref.~\cite{Hu:2003gh, Mortonson:2007hq} developed the principal component (PC) method, where a few PCs is sufficient to describe the entire model space of physical ionization histories regarding their observable impact on the large-angle $E$-mode power spectrum. This method has been applied to WMAP and Planck data to obtain complete constraints on reionization models~\cite{Mortonson:2008rx, Mortonson:2007hq, Heinrich:2016ojb, Aghanim:2018eyx}. It was also adopted for a Planck 2013 analysis for marginalizing ionization history when constraining inflationary parameters in Ref.~\cite{Planck:2013jfk}, as well as massive neutrinos and gravitational waves in Ref~\cite{Dai:2015dwa}\ch{[also cite our inflation papers?]}.

@@

Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood, called $\texttt{srollv2}$, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the steplike model is reduced by roughly a factor of two from \ch{XX to XX. [fill in numbers]}

@@

\section{PC Technique}
\label{sec:background}
The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}. We now briefly summarize the PC technique in \S\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in \S\ref{sec:KDE}. We refer the readers to Refs.~\cite{
%Hu:2003gh, %Mortonson:2008rx, 
Mortonson:2007hq,Heinrich:2016ojb} respectively for a more complete description of these topics.

@@

The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$. Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~(\ref{eq:xetommu}). Since $\mathbf{m}$ can take any continuous value, we approximate its effective likelihood with a kernel density estimate
\beq
{\cal L}_{\rm KDE}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
\eeq
where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$. To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx 1.1 \times 10^{6}$ chain samples. 
\ch{For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient for most models, and indeed we have tested it to work well for the ones we study in \S~\ref{sec:effective_likelihood}.}

@@

Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
\begin{align}
\tau(z,z_{\rm max}) & = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)}.
\label{eq:cumtau}
\end{align}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter. 
Given the tight constraints on cosmological parameters in the $\Lambda$CDM model, this quantity is well approximated by
\begin{align}
\tau_{\rm PC}(z,z_{\rm max})& = \sum_{a=1}^5 m_a \tau_a(z,z_{\rm max}) + \tau_{\rm fid}(z,z_{\rm max}),
\label{eq:cumtauPC}
\end{align}
where $\tau_a$ and $\tau_{\rm fid}$ are defined by employing Eq.~({\ref{eq:mmutoxe}}) in
Eq.~(\ref{eq:cumtau}) using the fiducial model 
for $H(z)$ and number densities.
 When we omit the redshift arguments, we
implicitly mean the total range, e.g. 
 $
 \tau_{\rm PC}=\tau_{\rm PC}(0,z_{\rm max}).
 $
 \ch{CH: what about recaling for different cosmological model? Our tau(z,zmax) plots have cosmological parameters marginalized, while Figs 9-10 (total tau posterio) do not.}
 \wh{They probably shouldn't, i.e should be constructed directly out of $m_a$, especially if the increased uncertainty is small.   If it is noticable we'll need to comment on this since our recommended approach is to just use $m_a$ (which has cosmology premarginalized)
 for other constraints and fix cosmology thereafter.}

@@

%
Note that the original $0\le x_e \le x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for a physical ionization history.  
In other words, no physical model is excluded by these priors but some unphysical models are included.  In \S \ref{sec:note_on_priors} we further explore the role of these and other choices of PC priors, and in \S\ref{sec:effective_likelihood} we test PC constraints against exact constraints for two example models.

@@

\begin{figure*}
%\includegraphics[width=0.7\textwidth]{plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML.png}
\includegraphics[width=0.8\textwidth]{paper/plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML_wGaussEllipse.pdf}
\caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$. We show  the 68\% and 95\% $m_a-m_b$ constraints as well as marginal constraints on individual $m_a$ for Planck 2018 in blue,
using \texttt{srollv2} for the low-$\ell$ $EE$ likelihood, and compare that to the Planck 2015 constraints in red.  The corresponding marginal distributions for individual $m_a$ are also shown.
%The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle.
When deriving model independent optical depth constraints we impose an additional physicality prior that excludes the gray region in $m_1$ and regions beyond the box limits in the others.
 The black solid line indicates the trajectory of tanh models with varying $\tau$. The red and black crosses indicate the best-fit tanh models from Planck 2015 and 2018 respectively. Note that the best-fit tanh model has moved to a lower $\tau$ in Planck 2018, and in particular, is now within the 68\% C.L. contours of the constraints.
}
\label{fig:plot_mjs_2018_vs_2015}
\end{figure*}
%
%
%
%
In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$. We also show the trajectory of the tanh model (see \S \ref{sec:example1}) through this space as well as the best fit tanh model points.
%
\footnote{The Planck 2015 best-fit for the tanh model is derived from a chain best-fit model as taken from Ref.~\cite{Heinrich:2016ojb}, whereas for the Planck 2018 point in this paper we used a minimizer to find the best-fit; there are also slight model differences, where in Planck 2018 we used a tanh width of $\Delta z = 0.015(1+z)$ instead of $\Delta z = 0.5$ for the Planck 2015 best-fit. Using $\Delta z = 0.5$ for Planck 2018 moves the cross negligibly on this plot.}
%
%
%
%
The gray shaded region in the $m_1-m_2$ plane represents the parameter region that would be excluded by an additional 
physicality prior from Eq.~(\ref{eq:individualprior}),
which is just beyond the border of the displayed regime for all but $m_1$. Results using the Planck 2018 data are shown in blue. While all five PC amplitudes are  constrained by Planck, the first two are particularly well-constrained. Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth, especially at high redshift.  
This preference also increases the impact of placing a physicality prior at low values of $m_1$ when deriving model-independent constraints as we shall discuss further in \S \ref{sec:modelindependent}.
We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}.
In the physicality prior excluded region (gray), the ionization and the cumulative optical depth at high redshift becomes negative.

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $m_1-m_2$ PC plane. We also 
the Planck 2018 constraints using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) and the \texttt{srollv2} likelihood (blue). The \texttt{srollv2} likelihood used better foreground removal and is what we use as our fiducial results throughout this paper. 
} \ch{keep same color scheme as Fig. 2, use green maybe for simallEE}
\label{fig:plot_m1m2_2015_vs_2018}
\end{figure}

In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we also show the constraints assuming a Gaussian posterior in $m_a$ using the PC chain covariance and means listed in Table~\ref{tab:PC_stats}.  The visual agreement with the full posteriors is remarkably good, especially in the well-constrained components.  This suggests that the simple Gaussian likelihood of Eq.~(\ref{eq:gaussian}) may suffice for rough constraints on models. \ch{We shall see in \S\ref{sec:effective_likelihood} that a quantitative comparison indeed shows good agreement between the KDE and Gaussian likelihoods, with only minor differences.}

@@

We now turn to constraining reionization models with PCs using the
effective likelihood approach described in~\S\ref{sec:effective_likelihood}.  \ch{The benefit of this approach is that
complete constraints on any model of the ionization history  between $6 \leq z \leq \zmax$ are obtained ahead of time and compressed into a simpler likelihood, so that when doing model testing one no longer needs to modify the $x_e(z)$ function in for example CAMB, each time a new ionization model needs to be tested. These provide savings on human time in addition to the computing speed-ups from the likelihood evaluations.}

\ch{Our code is available on GitHub at \url{...} [give URL]. 
The initial release corresponds to the fiducial results in this paper, i.e. derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{srollv2}$ likelihoods at $\zmax = 30$. 
The most important part of the code release is the python package called \texttt{Relike} which implements the Gaussian likelihood. The user can specify a customized $x_e(z)$ function taking in reionization model parameters and use \textt{Relike} to return the effective likelihood of the model. The returned likelihood is already marginalized over cosmological parameters. 
To obtain marginalized distributions more easily for high-dimensional parameter space, one may choose to sample \texttt{Relike} within a MCMC sampler such as Cobaya or Cosmosis \ch{[cite]}}

\ch{For convenience, we also provide a pre-assembled CosmoMC-Relike code using the CosmoMC sampler in generic mode, with which we have generated the results presented in this section. This CosmoMC-Relike code has an internal  implementation of the Gaussian and KDE likelihoods in fortran, which is separate from our main python package that only supports the Gaussian likelihood. }

\ch{When using the KDE likelihood, we suggest using the default value of $f = 0.14$ to avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation. All our KDE results in this section are computed with $f = 0.14$, which we shall soon show to work well in recovering results from an exact MCMC. Note that in both the Gaussian and the KDE modes, the model parameters or priors must be arranged to explicitly satisfy fully-ionized hydrogen and singly-ionized helium for $z\leq 6$. }

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]
{paper/plots/pl18_tau_posterior_tanh_vs_tanh_kde_vs_tanh_gaussian_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
\caption{Comparison of the total optical depth posterior for the tanh model in \refsec{example1} between the exact likelihood (shaded) and our effective likelihoods in the kernel density estimate (KDE) mode (blue solid) and the Gaussian approximation mode (black dashed). The KDE mode agree well with the exact results, while giving a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing at roughly $\sim$7\% level in the parameter posterior corresponding to $f = 0.14$. The Gaussian likelihood works well too, with no extra broadening of the posterior, but a negligible shift in the mean ($\sim XX\%$ of the error in $\tau$). This posterior is symmetric and does not capture the small skewness found in the exact and KDE chains. \ch{might need to rephrase.} 
}
\label{fig:tanh}

@@

\begin{figure}[ht]
\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_tanh_direct_vs_kde_vs_gaussian.pdf}
\caption{Comparison of the 68\% and 95\% C.L. constraints on the cumulative optical depth in the tanh model (described in \refsec{example1}) between the full Planck likelihoods (shaded bands), our effective likelihoods in the KDE mode (blue solid) and Gaussian mode (black dashed). The effective likelihoods results oscillate around zero at high-z because of the truncated representation of the models at 5 PCs (enough to represent any observable effects in the data). While the Gaussian approximation performs fairly well, the more complete KDE mode gives a slightly more accurate 95\% limits on the total optical depth. }
%double check plot, upper and lower limits seem to cross?
\label{fig:plot_taugtz_PC_vs_tanh}
\end{figure}

@@

\ch{The 95\% C.L. upper limit on high-redshift optical depth are also consistent: $\tau(15, 30) < 0.017$ for both effective likelihoods and $\tau(15, 30) < 0.016$ for the exact likelihood.}
\ch{Note also that the 95\% C.L. allowed two-step model in 
Fig.~\ref{fig:two_step_model} has
$\tau(15,30)= 0.013$.}
%0.01343375 evaluated at fiducial cosmology

>>>>

\title{\ch{\textsc{RELIKE}}: Reionization Effective Likelihood from Planck 2018 Data (?)}

@@

We publish \textsc{RELIKE} (Reionization Effective Likelihood), an fast and accurate effective likelihood code based on constraints from the latest Planck 2018 data on the CMB reionization principal components (PC). This code allows one to constrain arbitrary global ionization history $x_e(z)$ between $6 < z < 30$ with Planck without having to modify CAMB each time a different model is being tested. We tested the code on two example models which showed good agreement with sampling the exact Planck likelihoods.
We also find that contrary to Planck 2015, the Planck 2018 PC posteriors are well approximated by a Gaussian in PC amplitude space, this allows us to replace the kernel density estimate (KDE) evaluation of the PC chains in the code with an even faster Gaussian evaluation. We supply both the KDE and Gausisan modes of the likelihood in a CosmoMC implementation, and the Gaussian mode alone in an easy-to-use standalone python likelihood package that can be connected to any other sampler. We expect the code release to dramatically reduce the time for testing individual \ch{global} ionization models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as 21cm, luminosity function, star formation history, etc.
Finally, we derive model-independent constraints: $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ and $\tau_{\rm PC}(15, 30) < 0.023 $ (95\% C.L.). The upper limits on high-redshift optical depth is at least a factor of 2 looser than those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method; we demonstrate robustness of our results by explicitly recovering  \ch{$\tau_{\rm 2step}(15, 30) < 0.013 $ (95\% C.L.)} using a  two-step toy model without using PCs.}

@@

\ch{CH: going through the paper once from beginning to end now.}

\ch{CH: may also want to comment on Ref.~\cite{Paoletti:2020ndu} and other papers under paper/refs in overleaf (also just found \url{https://arxiv.org/pdf/2103.14378.pdf}.}

The cosmic microwave background (CMB) has entered an era of precision cosmology. The measurements from the Planck satellite have shown agreement with the standard $\Lambda$CDM model which describes the initial perturbations in the Universe and their evolution~\cite{Aghanim:2018eyx}. While many components of the standard cosmology model is well-understood, the details of the process of reionization however, remains one of the most uncertain pieces (see e.g.~\cite{2016ASSL..423.....M} and references therein). Its uncertainty propagates into the inferences of other important parameters such as the primordial power spectrum amplitude. Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB~ \cite{Smith:2006nk,Allison:2015qca}; and will also have implications for inferring cosmic acceleration through the growth of structure~\cite{Hu:2003pt}.

Typically, the impact of reionization on the primary CMB fluctuations has been modeled as a steplike transition in the global ionization history \ch{in the form of a tanh function}, with the step location parameterized by the total Thomson optical depth induced. This steplike \ch{tanh} model describes a Universe in which all hydrogen becomes fully ionized almost instantaneously at one particular redshift, and assumes, by construction, that there is negligible ionization before the transition. However, through the shape of the reionization bump induced in the CMB $E$-mode polarization at large angles, more information on the coarse-grained evolution of the ionization history can be obtained~\cite{Hu:2003gh, Mortonson:2007hq}.

In fact, to extract the most information possible from this $E$-mode bump, Ref.~\cite{Hu:2003gh, Mortonson:2007hq} developed the principal component (PC) method, where a few PCs is sufficient to describe the entire model space of physical ionization histories regarding their observable impact on the large-angle $E$-mode power spectrum. This method has been applied to WMAP and Planck data to obtain complete constraints on reionization models~\cite{Mortonson:2008rx, Mortonson:2007hq, Heinrich:2016ojb, Aghanim:2018eyx}. It was also adopted for a Planck 2013 analysis for marginalizing ionization history when constraining inflationary parameters in Ref.~\cite{Planck:2013jfk}, for studying inflationary features vs reionization features in Ref.~\cite{Obied:2018qdr}, as well as massive neutrinos and gravitational waves in Ref.~\cite{Dai:2015dwa}. %\cite{Obied:2017tpd}

@@

\ch{I'm here}
Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood,
$\texttt{srollv2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}
~\cite{Delouis:2019bub}, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{−0.0062}$~\cite{}.\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.

\todo{CH: might need to change all instances of \textt{srollv2} to \textt{SRoll2} as used in the original sroll2 paper; also possibly \textt{simall\_EE} to \textt{SimAll} as used in the Planck 2018 paper.}

@@

\section{PC Technique}
\label{sec:background}
The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}. Here, we  briefly summarize the PC technique itself in \S\ref{sec:PC}.   We present  techniques 
%kernel density estimate technique
for building its effective likelihood which can be used to rapidly explore any reionization model in \S\ref{sec:KDE}. We refer the readers to Refs.~\cite{
%Hu:2003gh, %Mortonson:2008rx, 
Mortonson:2007hq,Heinrich:2016ojb} respectively for a more complete description of these topics.

@@

The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$. Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~(\ref{eq:xetommu}). Since $\mathbf{m}$ can take any continuous value, we approximate its effective likelihood with a kernel density estimate
\beq
{\cal L}_{\rm KDE}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
\eeq
where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$. To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \sim 10^6$ chain samples. 
For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient for most models, and indeed we have tested it to work well for the ones we study in \S~\ref{sec:effective_likelihood}.

@@

Although reionization PCs are mainly a tool for model testing using complete information from large angle CMB polarization, as they do not reconstruct the rapidly varying aspects of $x_e(z)$ itself, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
\begin{align}
\tau(z,z_{\rm max}) & = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)}.
\label{eq:cumtau}
\end{align}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter. 
Given the tight constraints on cosmological parameters in the $\Lambda$CDM model, this quantity is well approximated by
\begin{align}
\tau_{\rm PC}(z,z_{\rm max})& = \sum_{a=1}^5 m_a \tau_a(z,z_{\rm max}) + \tau_{\rm fid}(z,z_{\rm max}),
\label{eq:cumtauPC}
\end{align}
where $\tau_a$ and $\tau_{\rm fid}$ are defined by employing Eq.~({\ref{eq:mmutoxe}}) in
Eq.~(\ref{eq:cumtau}) using the fiducial model 
for $H(z)$ and number densities.  This differs from Ref.~\cite{Heinrich:2016ojb} where their cosmological dependence was employed.
 When we omit the redshift arguments, we
implicitly mean the total range, e.g. 
 $
 \tau_{\rm PC}=\tau_{\rm PC}(0,z_{\rm max}).
 $

@@

%
Note that the original $0\le x_e \le x_e^{\rm max}$ condition cannot be strictly enforced here because we  keep only the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for a physical ionization history.  
In other words, no physical model is excluded by these priors but some unphysical models are included.  In \S \ref{sec:note_on_priors} we further explore the role of these and other choices of PC priors, and in \S\ref{sec:effective_likelihood} we test PC constraints against exact constraints for two example models.

@@

\begin{figure*}
%\includegraphics[width=0.7\textwidth]{plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML.png}
\includegraphics[width=0.8\textwidth]{paper/plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML_wGaussEllipse.pdf}
\caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$. We show  the 68\% and 95\% $m_a-m_b$ constraints as well as marginal constraints on individual $m_a$ for Planck 2018 in blue,
using \texttt{srollv2} for the low-$\ell$ $EE$ likelihood, and compare that to the Planck 2015 constraints in red.  The corresponding marginal distributions for individual $m_a$ are also shown.
%The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle.
When deriving model independent optical depth constraints we impose an additional physicality prior that excludes the gray region in $m_1$ and regions beyond the box limits in the others.
 The black solid line indicates the trajectory of tanh models with varying $\tau$. The red and black crosses indicate the best-fit tanh models from Planck 2015 and 2018 respectively. Note that the best-fit tanh model has moved to a lower $\tau$ in Planck 2018, and in particular, is now within the 68\% C.L. contours of the constraints.
}
\label{fig:plot_mjs_2018_vs_2015}
\end{figure*}
%
%
%
%
In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$. We also show the trajectory of the tanh model (see \S \ref{sec:example1}) through this space as well as the best fit tanh model points.\footnote{The Planck 2015 best-fit for the tanh model is derived from a chain best-fit model as taken from Ref.~\cite{Heinrich:2016ojb}, whereas for the Planck 2018 point in this paper we used a minimizer to find the best-fit; there are also slight model differences, where in Planck 2018 we used a tanh width of $\Delta z = 0.015(1+z)$ instead of $\Delta z = 0.5$ for the Planck 2015 best-fit. Using $\Delta z = 0.5$ for Planck 2018 moves the cross negligibly on this plot.}
%
%
%
%
The gray shaded region % in the $m_1-m_2$ plane 
represents the parameter space that would be excluded by an additional 
physicality prior from Eq.~(\ref{eq:individualprior}),
which is just beyond the border of the displayed regime for all but $m_1$. Results using the Planck 2018 data are shown in blue. While all five PC amplitudes are  constrained by Planck, the first two are particularly well-constrained. Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller $m_1$, leading to a reduced optical depth, especially at high redshift.  
This preference also increases the impact of placing a physicality prior at low values of $m_1$ when deriving model-independent constraints as we shall discuss further in \S \ref{sec:modelindependent}.
We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}.
In the physicality prior excluded region (gray), the ionization and the cumulative optical depth at high redshift becomes negative.

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (\wh{red}), Planck 2018 (blue, \texttt{srollv2}  \wh{decide whether to have this labeled}), and Planck 2018 but with the \texttt{simall\_EE} likelihood (\wh{green?}).   With Planck 2018, the \texttt{srollv2} likelihood moderately improves the best constrained direction and raises the optical depth.  }
\label{fig:plot_m1m2_2015_vs_2018}
\end{figure}

In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we also show the constraints assuming a Gaussian posterior in $m_a$ using the PC chain covariance and means listed in Table~\ref{tab:PC_stats}.  The visual agreement with the full posteriors is remarkably good, especially in the well-constrained components.  This suggests that the simple Gaussian likelihood of Eq.~(\ref{eq:gaussian}) may suffice for rough constraints on models. We shall see in \S\ref{sec:effective_likelihood} that a quantitative comparison indeed shows good agreement between the KDE and Gaussian likelihoods, with only minor differences in the shape of their posterior inferences.

@@

We now turn to constraining reionization models with PCs using the
effective likelihood approach described in~\S\ref{sec:effective_likelihood}.  The benefit of this approach is that
complete constraints on any model of the ionization history  between $6 \leq z \leq \zmax$ are obtained ahead of time and compressed into a simpler likelihood, so that when doing model testing one no longer needs to modify the $x_e(z)$ function \wh{and jointly sample the other cosmological parameters} in for example CAMB, each time a new ionization model needs to be tested. These provide savings on human time in addition to the computing speed-ups from the likelihood evaluations.

\ch{Our code is available on GitHub at \url{...} [give URL]. } The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{srollv2}$ likelihoods at $\zmax = 30$. 
The most important part of the code release is the python package called \texttt{Relike} which \wh{provides the PC parameters for a model and also} 
% added this since it now seemed to be emphasizing Gaussian only  - not sure that was intentional.
implements the Gaussian effective likelihood. The user can specify a customized $x_e(z)$ function taking in reionization model parameters and use \textt{Relike} to return the effective likelihood of the model. The returned likelihood is already marginalized over cosmological parameters. 
To obtain marginalized distributions more easily for high-dimensional parameter space, one may choose to sample \texttt{Relike} within a MCMC sampler such as Cobaya or Cosmosis \ch{[cite]}.

For convenience, 
\wh{query: since this is the place where KDE is given not sure ``convenience" is the transition we want here.   Do you want to say something about KDE up front?}
we also provide a pre-assembled CosmoMC-Relike code using the CosmoMC sampler in generic mode, with which we have generated the results presented in this section. This CosmoMC-Relike code has an internal  implementation of the Gaussian and KDE likelihoods in fortran, which is separate from our main python package that only supports the Gaussian likelihood.

When using the KDE likelihood, we suggest using the default value of $f = 0.14$ to avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation. All our KDE results in this section are computed with $f = 0.14$, which we shall soon show to work well in recovering results from an exact MCMC. Note that in both the Gaussian and the KDE modes, the model parameters or priors must be arranged to explicitly satisfy fully-ionized hydrogen and singly-ionized helium for $z\leq 6$.

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]
{paper/plots/pl18_tau_posterior_tanh_vs_tanh_kde_vs_tanh_gaussian_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
\caption{Total optical depth $\tau_{\rm tanh}$ posterior for the tanh model in \refsec{example1} for the exact (Planck 2018) likelihood (shaded), KDE PC 
(blue solid) and KDE Gaussian (black dashed)
effective likelihood. The overall agreement is excellent and the KDE results even capture the small skewness of the exact result. \wh{maybe $\tau_{\rm tanh}$ }
}
\label{fig:tanh}

@@

\begin{figure}[ht]
\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_tanh_direct_vs_kde_vs_gaussian.pdf}
\caption{Cumulative optical depth in the tanh model as in Fig.~\ref{fig:tanh}
(68\% and 95\% C.L. contours).  Overall agreement is good in regions where the contribution to $\tau_{\rm tanh}$ is high but the model form forbids high-$z$ contributions.   The KDE and Gaussian PC results  oscillate around zero at high-z because of effective low pass filter of the truncation to 5 PCs (enough to represent any observable effects in the data).  \wh{maybe $\tau_{\rm tanh}$} }
%double check plot, upper and lower limits seem to cross?
\label{fig:plot_taugtz_PC_vs_tanh}
\end{figure}

@@

\wh{Given the small amount of $\tau_{\rm hi}$ still allowed, the total optical depth constraints are correspondingly slightly larger than $\tau_{\rm tanh}$: give $\tau_{2\rm step}$.}
